A comparison of design-based and model-based approaches for finite population spatial sampling and inference.

### 2022-04-14 inst/manuscript/manuscript.Rmd
more minor updates

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference relies on random sampling. In the model-based approach, inference relies on distributional assumptions. We compare the approaches for finite population spatial data.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulated and real data. In the simulated and real data, a variety of sample sizes, location layouts, dependence structures, and response types are considered. The population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulated and real data, we found that regardless of the strength of spatial dependence in the data, the generalized random tessellation stratified (GRTS) algorithm, which explicitly incorporates spatial locations into sampling, tends to outperform the simple random sampling (SRS) algorithm, which does not explicitly incorporate spatial locations into sampling. We also found that model-based approaches tend to outperform design-based approaches, even for skewed data where the model-based distributional assumptions are violated. The performance gap between these approaches is small GRTS samples are used but large when SRS samples are used. This suggests that the sampling choice (whether to use GRTS or SRS) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial data that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhf{}
  \chead{Spatial design-based vs model-based}
  \doublespacing
output: rticles::elsevier_article
---

@@

When data cannot be collected for all units in a population (i.e., population units), data are collected on a subset of the population units -- this subset is called a sample. There are two general approaches for using samples to make frequentist statistical inferences about a population: design-based and model-based.  In the design-based approach, inference relies on randomly assigning some population units to be in the sample (random sampling). Alternatively, in the model-based approach, inference relies on distributional assumptions about the underlying stochastic process generating the sample. Each paradigm has a deep historical context [@sterba2009alternative] and its own set of benefits and drawbacks [@hansen1983evaluation, @brus1997random]. In this manuscript, we compare the design-based and model-based approaches for finite population spatial data.

@@

Spatial data are data that have some sort of spatial index, usually via coordinates. @de1990model and @brus1993design give early comparisons of design-based and model-based approaches for spatial data, quashing the belief that design-based approaches could not be used for spatially correlated data. Since then, there have been several general comparisons between design-based and model-based approaches for spatial data [@brus1997random; @verhoef2002sampling; @verhoef2008spatial; @brus2021statistical]. @cooper2006sampling reviews the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. In addition to @cooper2006sampling, there has been substantial research and development into estimators that use both design-based and model-based principles (see e.g., @sterba2009alternative and @cicchitelli2012model, and for Bayesian approaches, see @chan2020bayesian and @hofman2021many).

<!-- Paragraph about the contribution of the manuscript and outlining the rest -->
While comparisons between design-based and model-based approaches have been studied in spatial contexts, our contribution is comparing design-based approaches specifically built for spatial data to model-based approaches. Though the broad comparisons we draw between design-based and model-based approaches generalize to finite and infinite populations, we focus on finite populations.  A finite population contains a finite number of population units (we assume the finite number is known); an example is lakes (treated as a whole with the lake centroid representing location) in the conterminous United States.  An infinite population contains an infinite number of population units; an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we introduce and provide relevant background for design-based and model-based approaches to finite population spatial data. In Section \ref{sec:mm}, we describe how we intend to compare performance of the approaches using simulated and real data. In Section \ref{sec:results}, we present analysis reslts for the simulated and real data. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

@@

When samples are chosen in a manner such that the layout of sampled units reflects the layout of the population units, we call the resulting sample spatially balanced. By "reflecting the layout of the population units", we mean that if population units are concentrated in specific areas, the units in the sample should be concentrated in the same areas. Because spatially balanced samples reflect the layout of the population units, they are not necessarily spread out in space in some equidistant manner. One approach to selecting spatially balanced samples is the generalized random tessellation stratified (GRTS) algorithm [@stevens2004spatially], which we discuss in more detail in Section \ref{subsec:spb_design}.

@@

The model-based approach assumes the population is a random realization of a data-generating stochastic process (superpopulation). Randomness is formally incorporated through distributional assumptions on this process. Strictly speaking, randomness need not be incorporated through random sampling, though @diggle2010geostatistical warn against preferential sampling. Preferential sampling occurs when the process generating the data locations and the process being modeled are not independent of one another. To guard against preferential sampling, model-based approaches can implement some form of random sampling, though it is common for model-based approaches to sample non-randomly. When model-based approaches do implement random sampling, the inclusion probabilities are ignored when analyzing the sample (in contrast to the design-based approach, which relies on these inclusion probabilities to analyze the sample).

Instead of estimating fixed, unknown population parameters, as in the design-based approach, often the goal of model-based inference is to predict a realized variable. For example, suppose the realized mean of all population units (the realized population mean) is the variable of interest. Instead of a fixed, unknown mean, we are predicting the value of the mean, a random variable. Prediction intervals are then derived using assumptions of the data-generating stochastic process. If we repeatedly generate realizations from the same process and select samples, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide thorough reviews of model-based approaches for spatial data. In Fig. \ref{fig:fig1}, we provide a visual comparison of the design-based and model-based approaches (@verhoef2002sampling and @brus2021statistical provide similar figures). This figure contrasts the design-based approach with a fixed population and random sampling to the model-based approach with random populations and non-random sampling.

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples. Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples lacking spatial balance [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm to see widespread use was the generalized random tessellation stratified (GRTS) algorithm [@stevens2004spatially]. To quantify the spatial balance of a sample, @stevens2004spatially proposed loss metrics based on Voronoi polygons (i.e., Dirichlet Tessellations). After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, including stratified sampling with compact geographical strata @walvoort2010r, the local pivotal method [@grafstrom2012spatially; @grafstrom2018spatially], spatially correlated Poisson sampling [@grafstrom2012spatiallypoisson], balanced acceptance sampling [@robertson2013bas], within-sample-distance sampling [@benedetti2017spatially], and Halton iterative partitioning sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the GRTS algorithm because it is readily available in the `spsurvey` \textbf{\textsf{R}} package [@dumelle2022spsurvey] and naturally accommodates finite and infinite sampling frames, unequal inclusion probabilities, and replacement units. Replacement units are additional population units that can be sampled when a population unit originally selected can no longer be sampled. A couple reasons why an originally selected site can no longer be sampled include its location being physically inaccessible or on private land that the researcher does not have permission to access.

@@

FPBK assumes $\bm{\delta}$ in Equation$~$\ref{equation:Zmarginal} has mean-zero and a spatial dependence structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative, second-order stationary (depending only on the separation vector (e.g., distance) between population units), isotropic (independent of direction), and decay with distance between population units [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$, though @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or not either. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics]; one example is the exponential covariance function (@cressie1993statistics provides a thorough list of spatial covariance functions). The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter that quantifies the spatially dependent variability, $\sigma^2_{2}$ is the variance parameter the quantifies that spatially independent variability, $\phi$ is the distance parameter that measures the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. In geostatistical literature, $\sigma^2_{1}$ is often called the partial sill, $\sigma^2_{2}$ is often called the nugget, and $\phi$ is often called the range.

The parameters in Equation$~$\ref{equation:Zmarginal} can be estimated using a variety of techniques, but we focus on using restricted maximum likelihood [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing]. REML is preferred over maximum likelihood (ML) because ML estimates can be badly biased for small sample sizes, due to the fact that ML makes no adjustment for the simultaneous estimation of $\bm{\beta}$ and $\bm{\delta}$ [@patterson1971recovery].  Minus twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation}
where $\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s$ and $\bm{\Sigma}_{ss}$  is the covariance matrix of the sampled sites. Minimizing Equation$~$\ref{equation:reml} yields $\bm{\hat{\delta}}_{reml}$, the REML estimates of $\bm{\delta}$. Then $\bm{\beta}_{reml}$, the REML estimate of $\bm{\beta}$, is given by $(\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s$, where $\bm{\hat{\Sigma}}_{ss}$ is $\bm{\Sigma}_{ss}$ evaluated at $\bm{\hat{\delta}}_{reml}$.

With the model formulation in Equation$~$\ref{equation:Zmarginal}, the Best Linear Unbiased Predictor (BLUP) for $f(\mathbf{b}'\mathbf{z})$ and its prediction variance can be computed. While details of the derivation are in @verhoef2008spatial, we note here that the predictor and its variance are both moment-based, meaning that they do not rely on any distributional assumptions. Distributional assumptions are used, however, when constructing prediction intervals.

@@

In this section we describe how we used simulated and real data to investigate performance between simple random sampling without replacement (SRS) and GRTS sampling as well as performance between design-based (DB) and model-based (MB) inference. In SRS and GRTS sampling, all population units had equal inclusion probabilities. The important distinction between SRS and GRTS is that SRS ignores spatial locations while sampling but GRTS explicitly incorporates them. Together, the two sampling plans (SRS and GRTS) combined with the two inference approaches (DB and MB) yielded four sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB. For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to estimate means and the commonly-used SRS variance formula [@sarndal2003model; @lohr2009sampling] was used to estimate the variance. This variance formula is given by
\begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation}
where $z_i$ is the $i$th response value, $\bar{z}$ is the mean of all $z_i$, $n$ is the sample size, $N$ is the population size, and $f = (1 - n / N)$ ($f$ is often called the finite population correction factor). For GRTS-DB, the Horvitz-Thompson esetimator was used to estimate means and the local neighborhood variance was used to estimate variances. For SRS-MB and GRTS-MB, FPBK was used to estimate means and variances and parameters were estimated using restricted maximum likelihood.

We used simulated data to compare the sampling-inference combinations across many realized populations from the same data-generating stochastic process. With the simulated data, we were in control of the data-generating stochastic process and the random sampling process. We used real data from the 2012 National Lakes Assessment [@USEPA2012NLA] to compare the sampling-inference combinations within a single realized population (which is typically the case in reality). With the real data, we were in control of only the random sampling process.

@@

In each of the 36 simulation scenarios, there were 2000 independent simulation trials. Within each simulation scenario and trial, IRS and GRTS samples were selected and then design-based and model-based analyses were used to estimate (design-based) or predict (model-based) the mean and construct 95% confidence (design-based) or 95% prediction (model-based) intervals. With the model-based analyses, covariance parameters were estimated (using REML) separately for each trial.  After all 2000 trials, we summarized the long-run performance of the sampling-inference combination in each scenario by calculating mean bias, root-mean-squared error, and interval coverage. Mean bias is taken as the average deviation between each trial's estimated (or predicted) mean and its realized mean: $\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)$, where $i$ indexes simulation trials. Root-mean-squared error is taken as the square root of the average squared deviation between each trial's estimated (or predicted) mean and its realized mean: $\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)^2}$. Interval coverage is taken as the proportion of simulation trials where the realized mean is contained in its 95% confidence (or prediction) interval. These intervals are constructed using the normal distribution -- justification comes from the asymptotic normality of means via the central limit theorem (under some assumptions). Quantifying these metrics is important because together, they give us an idea of the accuracy (mean bias), spread (RMSE), and validity (interval coverage) of the sampling-inference combinations.

@@

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the contiguous United States. One component of NARS is the National Lakes Assessment (NLA), which measures various aspects of lake health and water quality. We focus on analyzing zooplankton multi-metric indices (ZMMI) and mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA. For ZMMI, data were collected at 1035 unique lakes. At less than 10% of lakes, two ZMMI replicates were collected. These were averaged for the purposes of our study so that each lake had one measurement for ZMMI. For Hg ppb, data were collected at 995 unique lakes (and there were no replicates like for ZMMI). The ZMMI and Hg ppb data are shown as spatial maps and as histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest near the coasts, lowest in the Central United States, higher near the coasts, are relatively symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest in the Northeastern United States, lowest elsewhere, are skewed, and have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram quantifies the the halved average squared differences (semivariance) of responses whose separation (distance) falls within some distance class. The spatial semivariance is closely related to the spatial covariance, and spatial semivariograms are often used to gauge the strength of spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance increases steadily with distance (meaning that observations nearby one another tend to be more similar than observations far apart from one another).

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial sampling and inference.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference relies on random sampling. In the model-based approach, inference relies on distributional assumptions. We compare the approaches for finite population spatial data.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulated and real data from the United States Environmental Protection Agency's 2012 National Lakes Assessment. A variety of sample sizes, location layouts, dependence structures, and response types are considered. The population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulated and real data, we found that regardless of the strength of spatial dependence in the data, the generalized random tessellation stratified (GRTS) algorithm, which explicitly incorporates spatial locations into sampling, tends to outperform the simple random sampling (SRS) algorithm, which does not explicitly incorporate spatial locations into sampling. We also found that model-based inference tends to outperform design-based inference, even for skewed data where the model-based distributional assumptions are violated. The performance gap between these design-based inference and model-based inference is small GRTS samples are used but large when SRS samples are used. This suggests that the sampling choice (whether to use GRTS or SRS) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial sampling and inference that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhf{}
  \chead{Spatial design-based vs model-based}
  \doublespacing
output: rticles::elsevier_article
---

@@

When data cannot be collected for all units in a population (i.e., population units), data are collected on a subset of the population units -- this subset is called a sample. There are two general approaches for using samples to make frequentist statistical inferences about a population: design-based and model-based.  In the design-based approach, inference relies on randomly assigning some population units to be in the sample (random sampling). Alternatively, in the model-based approach, inference relies on distributional assumptions about the underlying data-generating stochastic process (superpopulation). Each paradigm has a deep historical context [@sterba2009alternative] and its own set of benefits and drawbacks [@hansen1983evaluation; @brus1997random]. In this manuscript, we compare design-based and model-based approaches for finite population spatial sampling and inference.

@@

Spatial data are data that have some sort of spatial index --usually this index is specified via coordinates. @de1990model and @brus1993design give early comparisons of design-based and model-based approaches for spatial data, quashing the belief that design-based approaches could not be used for spatially correlated data. Since then, there have been several general comparisons between design-based and model-based approaches for spatial data [@brus1997random; @verhoef2002sampling; @verhoef2008spatial; @brus2021statistical]. @cooper2006sampling reviews the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. In addition to @cooper2006sampling, there has been substantial research and development into estimators that use both design-based and model-based principles (see e.g., @sterba2009alternative and @cicchitelli2012model, and for Bayesian approaches, see @chan2020bayesian and @hofman2021many).

<!-- Paragraph about the contribution of the manuscript and outlining the rest -->
While comparisons between design-based and model-based approaches have been studied in spatial contexts, our contribution is comparing design-based approaches specifically built for spatial data to model-based approaches. Though the broad comparisons we draw between design-based and model-based approaches generalize to finite and infinite populations, we focus on finite populations.  A finite population contains a finite number of population units (we assume the finite number is known) -- an example is lakes (treated as a whole with the lake centroid representing location) in the conterminous United States.  An infinite population contains an infinite number of population units -- an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we introduce and provide relevant background for design-based and model-based approaches to finite population spatial sampling and inference. In Section \ref{sec:mm}, we describe how we intend to compare performance of the approaches using simulated and real data. In Section \ref{sec:results}, we present analysis reslts for the simulated data and real data from the United States Environmental Protection Agency's 2012 National Lakes Assessment [@USEPA2012NLA]. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

@@

When samples are chosen in a manner such that the layout of sampled units reflects the layout of the population units, we call the resulting sample spatially balanced. By "reflecting the layout of the population units", we mean that if population units are concentrated in specific areas, the units in the sample should be concentrated in the same areas. Because spatially balanced samples reflect the layout of the population units, they are not necessarily spread out in space in some equidistant manner. One method of selecting spatially balanced samples is the generalized random tessellation stratified (GRTS) algorithm [@stevens2004spatially], which we discuss in more detail in Section \ref{subsec:spb_design}. To quantify the spatial balance of a sample, @stevens2004spatially proposed loss metrics based on Voronoi polygons (i.e., Dirichlet Tessellations).

@@

The model-based approach assumes the population is a random realization of a data-generating stochastic process. Randomness is formally incorporated through distributional assumptions on this process. Strictly speaking, randomness need not be incorporated through random sampling, though @diggle2010geostatistical warn against preferential sampling. Preferential sampling occurs when the process generating the data locations and the process being modeled are not independent of one another. To guard against preferential sampling, model-based approaches can implement some form of random sampling. It is common, however, for model-based approaches to sample non-randomly. When model-based approaches do implement random sampling, the inclusion probabilities are ignored when analyzing the sample (in contrast to the design-based approach, which relies on these inclusion probabilities to analyze the sample).

Instead of estimating fixed, unknown population parameters, as in the design-based approach, often the goal of model-based inference is to predict the value of a realized variable. For example, suppose the realized mean of all population units (the realized population mean) is the variable of interest. Instead of a fixed, unknown mean, we are predicting the value of the mean, a random variable. Prediction intervals are then derived using assumptions of the data-generating stochastic process. If we repeatedly generate realizations from the same process and select samples, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide thorough reviews of model-based approaches for spatial data. In Fig. \ref{fig:fig1}, we provide a visual comparison of the design-based and model-based approaches (@verhoef2002sampling and @brus2021statistical provide similar figures). Fig. \ref{fig:fig1} contrasts the design-based approach with a fixed population and random sampling to the model-based approach with random populations and non-random sampling.

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples. Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples lacking spatial balance [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm to see widespread use was the generalized random tessellation stratified (GRTS) algorithm [@stevens2004spatially]. After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, including stratified sampling with compact geographical strata @walvoort2010r, the local pivotal method [@grafstrom2012spatially; @grafstrom2018spatially], spatially correlated Poisson sampling [@grafstrom2012spatiallypoisson], balanced acceptance sampling [@robertson2013bas], within-sample-distance sampling [@benedetti2017spatially], and Halton iterative partitioning sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the GRTS algorithm because it is readily available in the `spsurvey` \textbf{\textsf{R}} package [@dumelle2022spsurvey] and naturally accommodates finite and infinite sampling frames, unequal inclusion probabilities, and replacement units. Replacement units are additional population units that can be sampled when a population unit originally selected can no longer be sampled. A couple reasons why an originally selected site can no longer be sampled include its location being physically inaccessible or on private land that the researcher does not have permission to access.

@@

FPBK assumes $\bm{\delta}$ in Equation$~$\ref{equation:Zmarginal} has mean-zero and a spatial dependence structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative, second-order stationary (depending only on the separation vector (e.g., distance) between population units), isotropic (independent of direction), and decay with distance between population units [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$. @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or not either. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics]; one example is the exponential covariance function (@cressie1993statistics provides a thorough list of spatial covariance functions). The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter that quantifies the spatially dependent variability, $\sigma^2_{2}$ is the variance parameter the quantifies that spatially independent variability, $\phi$ is the distance parameter that measures the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. In geostatistical literature, $\sigma^2_{1}$ is often called the partial sill, $\sigma^2_{2}$ is often called the nugget, and $\phi$ is often called the range.

The parameters in Equation$~$\ref{equation:Zmarginal} can be estimated using a variety of techniques, but we focus on using restricted maximum likelihood [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing]. REML is preferred over maximum likelihood (ML) because ML estimates can be badly biased for small sample sizes, due to the fact that ML makes no adjustment for the simultaneous estimation of $\bm{\beta}$ and $\bm{\delta}$ [@patterson1971recovery].  Minus twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation}
where $\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s$ and $\bm{\Sigma}_{ss}$  is the covariance matrix of the sampled sites. Minimizing Equation$~$\ref{equation:reml} yields $\bm{\hat{\delta}}_{reml}$, the REML estimates of $\bm{\delta}$. Then $\bm{\hat{\beta}}_{reml}$, the REML estimate of $\bm{\beta}$, is given by $(\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s$, where $\bm{\hat{\Sigma}}_{ss}$ is $\bm{\Sigma}_{ss}$ evaluated at $\bm{\hat{\delta}}_{reml}$.

With the model formulation in Equation$~$\ref{equation:Zmarginal}, the best linear unbiased predictor (BLUP) of $f(\mathbf{b}'\mathbf{z})$ and its prediction variance can be computed. While details of the derivation are in @verhoef2008spatial, we note here that the predictor and its variance are both moment-based, meaning that they do not rely on any distributional assumptions. Distributional assumptions are used, however, when constructing prediction intervals.

@@

In this section we describe how we used simulated and real data to investigate performance between simple random sampling (SRS) and GRTS sampling as well as performance between design-based (DB) and model-based (MB) inference. In SRS and GRTS sampling, all population units had equal inclusion probabilities and were selected without replacement. The important distinction between SRS and GRTS is that SRS ignores spatial locations while sampling but GRTS explicitly incorporates them. Together, the two sampling plans (SRS and GRTS) combined with the two inference approaches (DB and MB) yielded four sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB. For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to estimate means and the commonly-used SRS variance formula [@sarndal2003model; @lohr2009sampling] was used to estimate the variance. This variance formula is given by
\begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation}
where $z_i$ is the $i$th response value, $\bar{z}$ is the mean of all $z_i$, $n$ is the sample size, $N$ is the population size, and $f = (1 - n / N)$ ($f$ is often called the finite population correction factor). For GRTS-DB, the Horvitz-Thompson esetimator was used to estimate means and the local neighborhood variance was used to estimate variances. For SRS-MB and GRTS-MB, FPBK was used to estimate means and variances using restricted maximum likelihood.

We used simulated data to compare the sampling-inference combinations across many realized populations from the same data-generating stochastic process. With the simulated data, we were in control of the data-generating stochastic process and the random sampling process. We used real data from the 2012 National Lakes Assessment to compare the sampling-inference combinations within a single realized population (which is typically the case in reality). With the real data, we were in control of only the random sampling process.

@@

In each of the 36 simulation scenarios, there were 2000 independent simulation trials. Within each simulation scenario and trial, IRS and GRTS samples were selected and then design-based and model-based inferences were used to estimate (design-based) or predict (model-based) the mean and construct 95% confidence (design-based) or 95% prediction (model-based) intervals. With model-based inference, covariance parameters and the mean were estimated (using REML) separately for each trial.  After all 2000 trials, we summarized the long-run performance of the sampling-inference combination in each scenario by calculating mean bias, root-mean-squared error, and interval coverage. Mean bias is taken as the average deviation between each trial's estimated (or predicted) mean ($\hat{\mu}_i$) and its realized mean ($\mu_i$): $\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)$, where $i$ indexes simulation trials. Root-mean-squared error is taken as the square root of the average squared deviation between each trial's estimated (or predicted) mean and its realized mean: $\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)^2}$. Interval coverage is taken as the proportion of simulation trials where the realized mean is contained in its 95% confidence (or prediction) interval. These intervals are constructed using the normal distribution -- justification comes from the asymptotic normality of means via the central limit theorem (under some assumptions). Quantifying these metrics is important because together, they give us an idea of the accuracy (mean bias), spread (RMSE), and validity (interval coverage) of the sampling-inference combinations.

@@

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the contiguous United States. One component of NARS is the National Lakes Assessment (NLA), which measures various aspects of lake health and water quality. We focus on analyzing zooplankton multi-metric indices (ZMMI) and mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA. For ZMMI, data were collected at 1035 unique lakes. At less than 10% of lakes, two ZMMI replicates were collected. These were averaged for the purposes of our study so that each lake had one measurement for ZMMI. For Hg ppb, data were collected at 995 unique lakes (there were no replicates). The ZMMI and Hg ppb data are shown as spatial maps and as histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest near the coasts, lowest in the Central United States, are relatively symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest in the Northeastern United States, lowest elsewhere, are skewed, and have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram quantifies the the halved average squared differences (semivariance) of responses whose separation (distance) falls within some distance class. The spatial semivariance is closely related to the spatial covariance, and spatial semivariograms are often used to gauge the strength of spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance increases steadily with distance (meaning that observations nearby one another tend to be more similar than observations far apart from one another).
